---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---

by Chan Chee-Foong on 14 Apr 2016

***
### Libraries required
- lubridate
- dplyr
- ggplot2
- lattice
- knitr
- markdown
- rmarkdown

``` {r loadlibrary, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
## Install required packages if necessary

if(!is.element('lubridate', installed.packages()[,1])) {
    install.packages('lubridate')
}

if(!is.element('dplyr', installed.packages()[,1])) {
    install.packages('dplyr')
}

if(!is.element('ggplot2', installed.packages()[,1])) {
    install.packages('ggplot2')
}

if(!is.element('lattice', installed.packages()[,1])) {
    install.packages('lattice')
}


## Load the required libraries

library(lubridate)
library(dplyr)
library(ggplot2)
library(lattice)
```

***
## Loading and preprocessing the data

1. Download the zip file if it is not found in the working directory and unzip the file for analysis.

``` {r loaddata, echo=TRUE}
datadir <- "./activity"
datafile <- "activity.csv"
zipfile <- "activity.zip"

datadirfile <- paste(datadir, datafile, sep="/")
zipdirfile <- paste(datadir, zipfile, sep="/")

if (!file.exists(datadirfile)) {
    dir.create(datadir)
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url, destfile = zipdirfile, mode = 'wb')
    unzip (zipdirfile, exdir = datadir)
}
```

2. Load the csv file in to myData dataframe.  Update the date to the date class.  Update the interval to factor class.

```{r loaddata2, echo=TRUE}
myData <- read.csv(datadirfile, header = TRUE, na.strings = 'NA')
myData$date <- ymd(myData$date)
myData$interval <- as.factor(myData$interval)

```

3. Check data loaded

``` {r}
str(myData)
summary(myData)
head(myData)

```

***
## What is mean total number of steps taken per day?

Use dplyr library to group and summarise data.  Then calculate and show the mean and median.  

- *Notes: Could have used **aggregate** function here.  *


``` {r}
totalByDay <- myData[!is.na(myData$steps),] %>% group_by(date) %>% summarise(total = sum(steps))

head(totalByDay)

hist(totalByDay[!is.na(totalByDay$total),]$total, breaks = 5, col = 'blue', xlab = 'Total number of steps', main = 'Total number of steps taken per day')

myMean <- round(mean(totalByDay[!is.na(totalByDay$total),]$total),0)
myMedian <- median(totalByDay[!is.na(totalByDay$total),]$total)
abline(v=myMean, lwd = 2, lty= 2, col = 'green')
abline(v=myMedian, lwd = 2, lty= 3, col = 'yellow')

myMean
myMedian
```

Mean and median of total number of steps taken per day are <span style="color:red">**`r as.character(myMean)`**</span> and <span style="color:red">**`r as.character(myMedian)`**</span> respectively.

***
## What is the average daily activity pattern?

Use dplyr library to group and summarise data.  Then plot the mean number of steps every 5 minute interval throughout a day. 

``` {r}
meanByInterval <- myData[!is.na(myData$steps),] %>% group_by(interval) %>% summarise(mean = mean(steps))
meanByInterval <- meanByInterval[!is.na(meanByInterval$mean),]
meanByInterval$interval <- as.numeric(as.character(meanByInterval$interval))

head(meanByInterval)

ggplot(data=meanByInterval, aes(x=interval, y=mean)) + 
geom_line(size = 1, col = 'red') + 
xlab("5 minute interval in a day") + ylab("Number of steps") + 
ggtitle("Average number of steps at every 5 minute interval in a day")

myMax <- meanByInterval[meanByInterval$mean == max(meanByInterval$mean),]

myMax
```

The 5 minute interval of <span style="color:red">**`r as.character(myMax$interval)`**</span> contains the maximum mean number of steps of <span style="color:red">**`r as.character(round(myMax$mean,0))`**</span>


***
## Imputing missing values

Study the NA values by considering the following:  

- How many NA values are there?  
- On which days are there NA values?  
- Are these NA values occuring on days with values?  

``` {r}
countNA <- nrow(myData[is.na(myData$steps),])
countNA
unique(myData[is.na(myData$steps),]$date)
myData[!is.na(myData$steps) & myData$date %in% unique(myData[is.na(myData$steps),]$date),]
```

- There is a total number of <span style="color:red">**`r as.character(countNA)`**</span> rows with NAs in the dataset.  
- NA values occur only on 8 dates.
- On these 8 dates, all the steps values are NA.  

**Decision on imputing logic:**  
- The missing values are imputed based on the mean number of steps specific to weekday and interval.  
- The means are calculated based only on data with steps values.  

**Steps:**   
1. Mean number of steps specific to weekday and interval are calculated and populated in **meanByIntnDay**    
2. Merge with **myData** to form **myNewData**  
3. Populate the missing NA values  
4. Use dplyr library to group and summarise data.  
5. Plot the histogram

``` {r}
myData$day <- wday(myData$date, label=TRUE)
meanByIntnDay <- myData[!is.na(myData$steps),] %>% group_by(interval, day) %>% summarise(mean = mean(steps))
myNewData <- merge(myData, meanByIntnDay,by = c('interval','day'))
myNewData[is.na(myNewData$steps),]$steps <- myNewData[is.na(myNewData$steps),]$mean
myNewData$interval <- as.numeric(as.character(myNewData$interval))

head(myNewData)

newTotalByDay <- myNewData[!is.na(myNewData$steps),] %>% group_by(date) %>% summarise(total = sum(steps))

head(newTotalByDay)

hist(newTotalByDay[!is.na(newTotalByDay$total),]$total, breaks = 5, col = 'blue', xlab = 'Total number of steps', main = 'Total number of steps taken per day')

myNewMean <- round(mean(newTotalByDay[!is.na(newTotalByDay$total),]$total),0)
myNewMedian <- median(newTotalByDay[!is.na(newTotalByDay$total),]$total)

abline(v=myNewMean, lwd = 2, lty= 2, col = 'green')
abline(v=myNewMedian, lwd = 2, lty= 3, col = 'yellow')

myNewMean
myNewMedian
```

Mean and median of total number of steps taken per day are <span style="color:red">**`r as.character(myNewMean)`**</span> and <span style="color:red">**`r as.character(myNewMedian)`**</span> respectively.  
Before imputing the missing values, the mean and median are closed to each other.  After imputing the missing values, the mean and median are quite different as observed on the histograms.


***
## Are there differences in activity patterns between weekdays and weekends?

Additional field **daytype** was created to separate the data into 2 groups.  Sat and Sun defined as weekends and the rest of the days as weekdays.

``` {r}
myNewData$daytype <- as.factor(ifelse(myNewData$day %in% c('Sat','Sun'),'Weekend','Weekday'))

head(myNewData)

meanByIntnDT <- myNewData[!is.na(myNewData$steps),] %>% group_by(interval,daytype) %>% summarise(mean = mean(steps))

head(meanByIntnDT)

with(meanByIntnDT, 
    xyplot(mean ~ interval | daytype, type= 'l', 
    layout = c(1,2),
    lwd = 2, 
    main = 'Average number of steps taken in a day',
    xlab = '5 minute interval in a day',
    ylab = 'Number of steps',
    panel = function(x,y,...) {
        panel.xyplot(x,y,...)
        panel.abline(h=100, col='red')
    }
    ))

```

Noticed that there differences in activity patterns between weekdays and weekends.  For weekdays, most activities occur earlier part of the day.  For weekends, activities occur throughout the day.


***
### End of Assignment - Thank you for your time reviewing my work.  Have a nice day!
